The Timing and Likelihood of the Origin of Life from Earth's Impact History

### 2023-05-10 ms.tex
discussion

% Define document class
\documentclass[preprint]{aastex63}
\usepackage{tabularx}
\usepackage{amsmath}

@@

\section{Results}

\begin{figure}
  \centering
  \includegraphics[width=0.65\textwidth]{figures/example_impact_histories.pdf}
  \caption{Three simulated impact histories of the 5000 derived from our Monte-Carlo approach (Section \ref{sec:methods}). Each vertical line indicates an impact of a mass shown by the y-axis. The red lines indicate the last impact to vaporize the ocean assuming 10\% of an impacts kinetic energy heats the ocean.}
  \label{fig:example_impact_histories}
\end{figure}

Figure \ref{fig:example_impact_histories} illustrates three of the 5000 impact histories computed with our Monte-Carlo approach (Section \ref{sec:methods}). In Figure \ref{fig:example_impact_histories}a, a Moneta-sized impact occurs at $\sim 4.43$ Ga delivering nearly all of Earth's mantle HSEs. The impact should also reduce the Hadean atmosphere creating conditions favorable for prebiotic chemistry. However, in this case, Moneta is not a life-starting impact because Moneta is followed by an ocean-vaporizing collision at $\sim 4.37$ Ga which we assume destroys any existing biosphere. Based on smoothed-particle hydrodynamics impact simulations, we nominally assume that 10\% of an impact's kinetic energy goes into vaporizing the ocean \citep[Appendix \ref{sec:append_vap},][]{Citron_2022}. Vaporizing the ocean requires $5 \times 10^{27}$ J, so a $> 5 \times 10^{28}$ J impact should deliver sufficient energy under our assumptions. Appendix \ref{sec:append_vap} considers alternative criteria for ocean-vaporization which we discuss at the end of the Results section. In Figure \ref{fig:example_impact_histories}a, the last ocean-vaporizing impact is only $\sim 10^{20}$ kg (360 km) but has a large impact velocity of $32$ km/s giving it $5.12 \times 10^{28}$ J of kinetic energy.

@@

A caveat to the results above is that ocean-vaporization by an impact is likely a complicated function of impact mass, velocity, and incident angle, which we do not account for (Appendix \ref{sec:append_vap}). Through this section we have assumed that 10\% of an impactor's kinetic energy heats the planetary surface ($f_\mathrm{E,vap} = 0.1$), leading to a required $> 5 \times 10^{28}$ J collision to vaporize an ocean. Appendix Figure \ref{fig:energy_for_ocean_vap} uses simulations of impacts from \citet{Citron_2022} to show that $f_\mathrm{E,vap} = 0.1$ is a reasonable assumption, but that values between 2.5\% and 25\% are also possible and that $f_\mathrm{E,vap}$ likely also depends on impact angle.

@@

\section{Discussion and Conclusions}

@@

We emphasize that these results rely upon a variety of assumptions including the premise that some of the first life-like molecules were RNA replicators which were built abiotically from nitriles like HCN and HCCCN. Additionally, our calculations assume that an impact that vaporizes the ocean also sterilizes the planet, but this may not be the case because microbes could have possibly survived in the deep subsurface \citep{Sleep_1989,Grimm_2018}.

If life began in an RNA-first process after an impact, then this work suggests that origin of life on Earth was a fairly likely event (at least a 20\% chance) and was indeed strongly favored (92\% chance) under optimistic assumptions. Given that all rocky planets form from accretionary impacts, our work supports and optimistic outlook for life on exoplanets and future searches for biosignatures.

>>>>

% Define document class
\documentclass[manuscript]{aastex63}
\usepackage{tabularx}
\usepackage{amsmath}

@@

\section{Results} \label{sec:results}

\begin{figure}
  \centering
  \includegraphics[width=0.65\textwidth]{figures/example_impact_histories.pdf}
  \caption{Three simulated impact histories of the 5000 derived from our Monte-Carlo approach (Section \ref{sec:methods}). Each vertical line indicates an impact of a mass shown by the y-axis. The red lines indicate the last impact to vaporize the ocean assuming 10\% of an impacts kinetic energy heats the ocean. (a) gives conversions between impact mass and diameter for 500, 1000 and 2000 km asteroids.}
  \label{fig:example_impact_histories}
\end{figure}

Figure \ref{fig:example_impact_histories} illustrates three of the 5000 impact histories computed with our Monte-Carlo approach (Section \ref{sec:methods}). In Figure \ref{fig:example_impact_histories}a, a Moneta-sized impact occurs at $\sim 4.43$ Ga delivering nearly all of Earth's mantle HSEs. The impact should also reduce the Hadean atmosphere creating conditions favorable for prebiotic chemistry. However, in this case, Moneta is not a life-starting impact because Moneta is followed by an ocean-vaporizing collision at $\sim 4.37$ Ga which we assume destroys any existing biosphere. Based on smoothed-particle hydrodynamics impact simulations, we nominally assume that 10\% of an impact's kinetic energy goes into vaporizing the ocean \citep[Appendix \ref{sec:append_vap},][]{Citron_2022}. Vaporizing the ocean requires $5 \times 10^{27}$ J, so a $> 5 \times 10^{28}$ J impact should deliver sufficient energy under our assumptions. Appendix \ref{sec:append_vap} considers alternative criteria for ocean-vaporization. In Figure \ref{fig:example_impact_histories}a, the last ocean-vaporizing impact is only $\sim 10^{20}$ kg (360 km) but has a large impact velocity of $32$ km/s giving it $5.12 \times 10^{28}$ J of kinetic energy.

@@

\section{Discussion}

Our estimated timing for biopoesis is compatible with the earliest well-accepted geologic evidence of life in the 3.5 Ga Pilbara block of western Australia \citep{Walter_1980}. Older geologic evidence of life exists, but is inconclusive. For example, \citet{Bell_2015} discovered graphite inclusions in a 4.1 Ga zircon grain with $^{13}$C depletions compatible with biological activity, but this is a tentative biosignature because there are abiotic mechanisms for making isotopically light carbon \citep{Javaux_2019}. If the \citet{Bell_2015} finding is indeed a sign of life, then it would be mostly compatible with our estimating timing in Figure \ref{fig:time_and_mass}, because we find that there is only a $\sim 10\%$ chance of life starting after 4.1 Ga.

Beyond the early Earth, it is temping to extrapolate our result for the likelihood of the origin of life to exoplanets, however, we must do so with caution because planet's orbiting different stars may have bombardment histories unlike the Hadean. Planets in the habitable zone of an M-type star during the stellar main-sequence phase were interior to the habitable zone for several hundred million years during the super luminous pre-main-sequence phase \citep{Luger_2015}. \citet{Lichtenberg_2022} uses N-body simulations to show that, for planets hosted by M-dwarfs, large asteroid impacts are not useful for prebiotic chemistry because big impacts most likely occur within 100 million years of planet formation during the stellar pre-main-sequence when the planet is outside of the habitable zone. The pre-main-sequence phase is not an issue for an impact-induced origin of life on planets orbiting sun-like stars because the phase only lasts 10s Myrs, while large asteroid impacts occur for 100s Myrs \citep{Lichtenberg_2022}.

A caveat to our results is that ocean-vaporization by an impact is likely a complicated function of impact mass, velocity, and incident angle, which we do not account for (Appendix \ref{sec:append_vap}). Throughout Section \ref{sec:results} we have assumed that 10\% of an impactor's kinetic energy heats the planetary surface ($f_\mathrm{E,vap} = 0.1$), leading to a required $> 5 \times 10^{28}$ J collision to vaporize an ocean. Appendix Figure \ref{fig:energy_for_ocean_vap} uses simulations of impacts from \citet{Citron_2022} to show that $f_\mathrm{E,vap} = 0.1$ is a reasonable assumption, but that values between 2.5\% and 25\% are also possible and that $f_\mathrm{E,vap}$ likely also depends on impact angle.

@@

An additional related caveat is that our calculations assume that an impact that vaporizes the ocean also sterilizes the planet, but this may not be the case because microbes could have possibly survived in the deep subsurface \citep{Sleep_1989,Grimm_2018}. If ocean-vaporization does not destroy the biosphere, then our results would favor an origin of life earlier than we predict in Figure \ref{fig:time_and_mass}, and a life-starting impact would also be more probable than we have estimated (Figure \ref{fig:probabilities_of_impacts}).

\section{Conclusions}

@@

If life began in an RNA-first process after an impact, then this work suggests that origin of life on Earth was a fairly likely event (at least a 20\% chance) and was indeed strongly favored (92\% chance) under optimistic assumptions. Given that rocky planets form from accretionary impacts, our work supports and optimistic outlook for life on exoplanets orbiting sun-like stars and future searches for biosignatures.
